---
title: "AEMET OpenData to Functional Data Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AEMET OpenData to Functional Data Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

# Overview

This vignette illustrates a complete workflow to download daily meteorological
data from **AEMET OpenData** and convert them into functional data objects of
the **fda.usc** package using **aemet2fdata**.

- AEMET OpenData portal: https://opendata.aemet.es/centrodedescargas/inicio
- API key request: https://opendata.aemet.es/centrodedescargas/altaUsuario

Every functional object has **one curve per station and year** with 365
points (`argvals = 1:365`). February 29 is removed, so column 60 is always
March 1. Missing days are kept as `NA`.

# 1. Example without API key

The package ships daily data of two stations (1387, A Coruña, and B228,
Palma Puerto) for 2023-2024. The same code works with a file saved by
`aemet2df(..., file = "mydata.csv")`.

```{r setup}
library(aemet2fdata)
library(fda.usc)
f <- system.file("extdata", "aemet_example.csv", package = "aemet2fdata")
```

## One variable: `fdata`

```{r fdata}
fd_tmed <- aemet2fdata(file = f, var = "tmed")
dim(fd_tmed)
rownames(fd_tmed$data)
```

```{r fdata-plot}
plot(fd_tmed, col = c( 2, 2, 4, 4))
legend("topleft", legend = c("1387", "B228"), col = c(2, 4), lty = 1)
```

The usual **fda.usc** tools apply directly:

```{r fdata-mean}
plot(func.mean(fd_tmed), main = "Mean curve", lwd = 2)
```

## Several variables: `ldata`

`aemet2lfdata()` returns an `ldata` object: `df` has one row per curve
(station-year) with the station metadata, and there is one `fdata` per
variable, all with the same rows.

```{r ldata}
ld <- aemet2lfdata(file = f, vars = c("tmed", "tmax", "tmin", "prec"))
names(ld)
sapply(ld, NROW)
ld$df
is.ldata(ld)
```

Subsets and plots use the `ldata` methods of **fda.usc**:

```{r ldata-subset}
ld_coruna <- subset(ld, ld$df$station_id == "1387")
sapply(ld_coruna, NROW)
```

```{r ldata-plot}
plot(ld$tmax, col = as.integer(factor(ld$df$station_id))*2,
     main = "Maximum temperature")
```

# 2. Downloading from the API

```{r api-key, eval=FALSE}
api_key <- Sys.getenv("AEMET_API_KEY")
```

## Station inventory

`aemet2inventory()` returns one row per station with `station_id`,
`station_name`, `province`, `altitude` (m), `lon`, `lat` (decimal degrees) and
`wmo_id`. Save it once and reuse it: the daily data do not include
coordinates, so the inventory is what provides `lon` and `lat` in
`ldata$df`.

```{r inventory, eval=FALSE}
inv <- aemet2inventory(api_key = api_key, file = "inventory.rds")
head(inv)
```

## Daily data

```{r example-df, eval=FALSE}
df <- aemet2df(
  station_ids = c("1387", "B228"),
  start_date = "2018-01-01",
  end_date = "2020-12-31",
  vars = c("tmed", "prec", "tmin", "tmax"),
  api_key = api_key,
  file = "mydata.csv"
)
head(df)
```

## Functional objects

```{r example-ldata, eval=FALSE}
ld <- aemet2lfdata(file = "mydata.csv", inventory = "inventory.rds")
ld$df
plot(ld)

# or directly from the API
ld <- aemet2lfdata(
  station_ids = c("1387", "B228"),
  vars = c("tmed", "prec"),
  start_date = "2018-01-01",
  end_date = "2020-12-31",
  api_key = api_key,
  inventory = inv
)
```

## One station, whole history: `aemet2csv()`

`aemet2csv()` requests each station in half-year windows (1 January-30 June
and 1 July-31 December, i.e. two requests per year), from the most recent
backwards, and writes **one CSV per station** with its whole daily history.
The CSV is written in UTF-8 whatever the locale, so station names with
accents or "ñ" are kept. Stations already downloaded are skipped, so an
interrupted run is resumed by running the same call again.

```{r one-station, eval=FALSE}
log <- aemet2csv(c("1387"), start_date = "2006-01-01",
                 end_date = "2025-12-31", api_key = api_key,
                 out_dir = "aemet_csv")
log

ld <- aemet2lfdata(file = "aemet_csv/1387.csv",
                   vars = c("tmed", "tmin", "tmax", "prec", "sol"),
                   inventory = "inventory.rds")
ld <- subset(ld, ld$df$n_days >= 365)          # complete years
plot(ld$tmed, col = hcl.colors(nrow(ld$df), "Blue-Red 3"))
```

## All stations over long periods

For many stations and years use `aemet2download()`: it requests all stations
at once (AEMET `todasestaciones` service) in short windows and, when a year is
complete, saves **one file per station and year**
(`aemet_daily/1387/1387_1976.rds`, ...). It is resumable: run the same call
again to continue or to retry failed windows. Then pass the directory to
`aemet2lfdata()` (optionally with `station_ids` to read only some stations).

```{r bulk, eval=FALSE}
log <- aemet2download(
  start_date = "2006-01-01",
  end_date   = "2025-12-31",
  api_key    = api_key,
  out_dir    = "aemet_daily",
  vars       = c("tmed", "tmin", "tmax", "prec", "sol")
)
table(log$status)

ld <- aemet2lfdata(file = "aemet_daily", inventory = "inventory.rds")
table(ld$df$year)

# complete years (>= 360 days) of stations with the 50 years
ld  <- subset(ld, ld$df$n_days >= 360)
ny  <- table(ld$df$station_id)
ld50 <- subset(ld, ld$df$station_id %in% names(ny)[ny == 50])
```

Station identifiers are `station_id` (AEMET *indicativo*, e.g. `"1387"`),
not `wmo_id`.

# 3. Built-in dataset from fda.usc

**fda.usc** includes the dataset `aemet` (73 stations, averages 1980-2009):

```{r aemet-data}
data(aemet)
names(aemet)
plot(aemet$temp, main = "fda.usc::aemet, mean temperature 1980-2009")
```

It is a **climatological** (averaged) version, with one curve per station,
whereas **aemet2fdata** gives one curve per station and year.

# 4. Recommended workflow for large downloads

1. Download the inventory once: `aemet2inventory(api_key, file = "inventory.rds")`.
2. Download data once: `aemet2csv()` (some stations, whole history),
   `aemet2download()` (all stations) or `aemet2df(..., file = "mydata.csv")`.
3. Build functional objects from the files:
   `aemet2lfdata(file = "mydata.csv", inventory = "inventory.rds")`.

This avoids repeated API calls and respects the AEMET OpenData service.

# Related packages

**climaemet** and **meteospain** also access AEMET OpenData: the first covers
many AEMET services with tidy/spatial output and climate graphics, the second
several Spanish meteorological services with a common format.
**aemet2fdata** focuses on long daily series stored locally and on their
conversion into functional data for **fda.usc**. Data source: © AEMET.
